---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Reproducible Research: Peer Assessment 1
========================================
Harry Braviner


## Loading and preprocessing the data

We unzip the data (if necessary), read it in as a data frame and convert the **date** variable from the *factor* to the *date* datatype.

```{r echo=TRUE}
if(!file.exists("activity.csv"))
  unzip("activity.zip")
data <- read.csv("activity.csv")
data$date <- as.Date(data[,"date"])
```



## What is mean total number of steps taken per day?

Let's first compute a vector containing the total number of steps taken each day and produce a histogram (I'm putting a rug under it since I think it adds useful information).

```{r, echo=TRUE, fig.width=7, fig.height=7}
stepCounts <- by(data$steps, data$date, sum, na.rm=TRUE)
hist(stepCounts, breaks=10, col = "blue", main = "", xlab = "Total number of steps taken in a day")
rug(stepCounts)
```

Now we compute the mean and median steps-per-day.

```{r, echo=TRUE}
meanSteps <- mean(stepCounts)
medianSteps <- median(stepCounts)
```

The mean number of steps-per-day is `r meanSteps` and the median number of steps-per-day is `r medianSteps`.

Note that this method has treated the NAs are zeros. Only averaging over the days for which all the data is available would give a different mean and median.

## What is the average daily activity pattern?

Create a vector of the averages over all days of the number of steps in each interval.
The names of this vector correspond to the times.

```{r, echo=TRUE}
averagedStepsPerInterval <- by(data$steps, data$interval, mean, na.rm=TRUE, simplify=TRUE)
```

I've changed the labelling of the x-axis to be the actual time, since the interval numbers aren't very nicely formatted (i.e. I think "15:00" is clearer than "1500" as a label for 3pm!)

```{r, echo=TRUE, fig.width=7, fig.height=5}
plot(names(averagedStepsPerInterval), averagedStepsPerInterval, type="l", col="red",
      xlab="Time", ylab="Steps in 5 minute interval", xaxt="n",
      main="Steps per interval averaged over all days")
axis(side=1, at = c("0", "600", "1200", "1800", "2355"),
      labels = c("00:00", "06:00", "12:00", "18:00", "23:55"))
```

Calculate interval with the maximum steps:

```{r, echo = TRUE}
maxStepInterval <- data$interval[which.max(averagedStepsPerInterval)[[1]]]
```

The five-minute interval containing the maximum number of steps is `r maxStepInterval`.

## Imputing missing values

```{r, echo=TRUE}
totalMissing <- sum(is.na(data$steps) | is.na(data$date) | is.na(data$interval))
totalMissing
```
So there are `r totalMissing` observations where at least one of the variables is missing.

As an aside, we can see from the next bit of code that all of these NAs are in the **steps** variables;
no observations of the **date** or **interval** have been coded as NA.
```{r, echo=TRUE}
sum(is.na(data$date))
sum(is.na(data$interval))
```

My strategy for filling in the missing data is to replace each missing value with the mean for that 5-minute interval, averaged over all the days for which that interval does have data.
This is convenient, since I computed a vector of such values in the previous section.

```{r, echo=TRUE}
# Note that this makes a copy of the data frame
# (unlike a data table, where we'd just create a reference)
dataWithMeans <- data
# Now loop through the rows replacing the steps entry with the mean
# value as necessary
for (i in 1:nrow(dataWithMeans)){
  if(is.na(dataWithMeans[i,"steps"]))
    dataWithMeans[i,"steps"] <- averagedStepsPerInterval[as.character(dataWithMeans[i,"interval"])]
}
```

Just as a check that this has been done correctly, let's look at a couple of examples where the original data has missing values and check that the replacement has worked

```{r, echo=TRUE}
data[10, "steps"]
dataWithMeans[10, "steps"]
dataWithMeans[10, "interval"]
averagedStepsPerInterval["45"]
```
So this has indeed replaced NAs with values from the vector of means.

We compute a histogram of the total number of steps taken each day, using this new data with imputed values:
```{r, echo=TRUE, fig.width=7, fig.height=7}
stepCountsWithMeans <- by(dataWithMeans$steps, dataWithMeans$date, sum, na.rm=TRUE)
hist(stepCountsWithMeans, breaks=10, col = "blue", main = "", xlab = "Total number of steps taken in a day")
rug(stepCountsWithMeans)
```

Now we compute the mean and median steps-per-day, using the data with imputed values:

```{r, echo=TRUE}
meanStepsWithMeans <- mean(stepCountsWithMeans)
medianStepsWithMeans <- median(stepCountsWithMeans)
```

The mean number of steps-per-day is `r format(meanStepsWithMeans, scientific=F)` and the median number of steps-per-day is `r format(medianStepsWithMeans, scientific = F)`.
These both differ from the values we found from the data ignoring the NAs, and are the same as each other!
This isn't surprising: the values with which we've filled in the blanks are the means for *that five-minute interval*, which will change the mean number of steps-per-day.
A lot of days have entirely missing data, and these are filled in with `r format(meanStepsWithMeans, scientific=F)` steps from the five-minute averages.
This value appears a lot, and so it isn't surprising that the median falls in here.

## Are there differences in activity patterns between weekdays and weekends?

First let's add a new variable called **weekday** to the data, using an anonymous function.
```{r, echo=TRUE}
dataWithMeans$weekday <- sapply(dataWithMeans$date,
                                 function(d) {w <- weekdays(d);
				               if(w == "Saturday" | w == "Sunday") {
					         return("weekend")
					       } else {
					         return("weekday")
					       }
					      })
#The new variable is of type character, so we now convert it to a factor variable
dataWithMeans$weekday <- as.factor(dataWithMeans$weekday)
```

Now create a pair of vectors containing the mean across all weekdays and weekend days.
```{r, echo=TRUE}
weekdayMeanSteps <- by(dataWithMeans[dataWithMeans$weekday == "weekday", "steps"],
                        dataWithMeans[dataWithMeans$weekday == "weekday", "interval"],
			mean)
weekendMeanSteps <- by(dataWithMeans[dataWithMeans$weekday == "weekend", "steps"],
                        dataWithMeans[dataWithMeans$weekday == "weekend", "interval"],
			mean)
```

And finally we create our two plots.
I'm going to use the base (rather than the lattice) plotting system because that's what I know best at this point.
```{r, echo=TRUE}
par(mfcol=c(2,1), mar = c(0,4,0,2), oma=c(5,0,2,0))
plot(names(weekendMeanSteps), weekendMeanSteps, type="l", col="blue", lwd=2,
      xlab = "", ylab="Steps in 5 minute interval", xaxt="n", main="")
axis(side=1, at=c("0000", "600", "1200", "1800", "2355"),
      #labels=c("00:00", "06:00", "12:00", "18:00", "23:55"))
      labels=c("", "", "", "", ""))
# Plot some vertical guide lines just to make the plot clearer
abline(v = c("600", "1200", "1800"), lty=2, lwd=0.5)
# Label this as the weekend data
text("0300", 150, "Weekend", font=2)
plot(names(weekdayMeanSteps), weekdayMeanSteps, type="l", col="red", lwd=2,
      xlab = "", ylab="Steps in 5 minute interval", xaxt="n", main="")
axis(side=1, at=c("0000", "600", "1200", "1800", "2355"),
      labels=c("00:00", "06:00", "12:00", "18:00", "23:55"))
# Plot some vertical guide lines just to make the plot clearer
abline(v = c("600", "1200", "1800"), lty=2, lwd=0.5)
# Label this as the weekday data
text("0300", 200, "Weekday", font=2)
# Setting xlab="Time" in the above plot() call doesn't seem to actually
# give me a label on the x-axis - probably because it has zero margin
# size to draw in. The following gives us the x-axis label
mtext("Time", side=1, outer=TRUE, line=3)
```
