---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Reproducible Research: Peer Assessment 1
========================================
Harry Braviner


## Loading and preprocessing the data

We unzip the data (if necessary), read it in as a data frame and convert the **date** variable from the *factor* to the *date* datatype.

```{r echo=TRUE}
if(!file.exists("activity.csv"))
  unzip("activity.csv")
data <- read.csv("activity.csv")
data$date <- as.Date(data[,"date"])
```



## What is mean total number of steps taken per day?

We first compute a vector containing the total number of steps taken each day and produce a histogram (I'm putting a rug under it since I think it adds useful information).

```{r, echo=TRUE, fig.width=7, fig.height=7}
stepCounts <- by(data$steps, data$date, sum, na.rm=TRUE)
hist(stepCounts, breaks=10, col = "blue", main = "", xlab = "Total number of steps taken in a day")
rug(stepCounts)
```

Now we compute the mean and median steps-per-day.

```{r, echo=TRUE}
meanSteps <- mean(stepCounts)
medianSteps <- median(stepCounts)
```

The mean number of steps-per-day is `r meanSteps` and the median number of steps-per-day is `r medianSteps`.

## What is the average daily activity pattern?

Create a vector of the averages over all days of the number of steps in each interval.
The names of this vector correspond to the times.

```{r, echo=TRUE}
averagedStepsPerInterval <- by(data$steps, data$interval, mean, na.rm=TRUE, simplify=TRUE)
```

I've changed the labelling of the x-axis to be the actual time, since the interval numbers aren't very nicely formatted (i.e. I think "15:00" is clearer than "1500" as a label for 3pm!)

```{r, echo=TRUE, fig.width=7, fig.height=5}
plot(names(averagedStepsPerInterval), averagedStepsPerInterval, type="l", col="red",
      xlab="Time", ylab="Steps in 5 minute interval", xaxt="n",
      main="Steps per interval averaged over all days")
axis(side=1, at = c("0", "600", "1200", "1800", "2355"),
      labels = c("00:00", "06:00", "12:00", "18:00", "23:55"))
```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
